A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950

### 2022-02-08 getdata4.Rmd
fix errors and remove socStatus

```{r fromobs}
bobs <- obs[obs$starttyp == 2 & obs$id %in% per$id, ]
indx <- match(per$id, bobs$id)
per$socStatus <- bobs$socStatus[indx]
per$socBranch <- bobs$socBranch[indx]
per$hisclass <- bobs$hisclass[indx]
per$nofrs <- bobs$nofrs[indx]
per <- per[!is.na(per$nofrs), ]
per$subreg <- factor(per$nofrs)
levels(per$subreg) <- c(rep("ume", 2), 
                        rep("bjur", 4), 
                        rep("sten", 3))
##per <- per[(per$subreg != "sten"), ] # NOTE: Now Stensele station is included
per$subreg <- factor(per$subreg)
per$enter <- 0
per$sex <- factor(per$kon, labels = c("boy", "girl"))
per$birthdate <- toTime(per$birthdate)
per$illeg <- as.numeric(per$ab == 2)
per$parity <- per$paritet_g
per$paritet_g <- per$paritet_k <- per$doddat <- per$frsbosdat <- NULL
per$fid <- per$fboort <- per$ab <- per$fyrkfd <- per$fyrkind <- NULL
per$region <- per$kon <- per$fodhfrs <- per$frsbobdat <- per$frsbostyp <- NULL
per$fbonmn <- per$fboind <- per$fyrkrel <- per$fyrketxt <- NULL
per$foddat <- as.Date(as.character(per$foddat), format = "%Y%m%d")
## NEW:
per$socBranch <- as.character(per$socBranch)
per$socBranch[is.na(per$socBranch)] <- "none"
per$socBranch <- factor(per$socBranch)
per$socStatus <- as.character(per$socStatus)
per$socStatus[is.na(per$socStatus)] <- "unknown"
per$socStatus <- factor(per$socStatus)
per <- per[, c("id", "sex", "birthdate", "mid", "foddat", "parity", "illeg", 
               "subreg", "socBranch", "socStatus", "hisclass", "enter", "exit", "event")]
#saveRDS(per, file = "data/infdat3.rds")
```

>>>>

```{r fromobs}
bobs <- obs[obs$starttyp == 2 & obs$id %in% per$id, ]
indx <- match(per$id, bobs$id)
per$socStatus <- bobs$socStatus[indx]
per$socBranch <- bobs$socBranch[indx]
per$hisclass <- bobs$hisclass[indx]
per$nofrs <- bobs$nofrs[indx]
per <- per[!is.na(per$nofrs), ]
per$subreg <- factor(per$nofrs)
levels(per$subreg) <- c(rep("ume", 2), 
                        rep("bjur", 4), 
                        rep("sten", 3))
##per <- per[(per$subreg != "sten"), ] # NOTE: Now Stensele station is included
per$subreg <- factor(per$subreg)
per$enter <- 0
per$sex <- factor(per$kon, labels = c("boy", "girl"))
per$birthdate <- toTime(per$birthdate)
per$illeg <- as.numeric(per$ab == 2)
per$parity <- per$paritet_g
per$paritet_g <- per$paritet_k <- per$doddat <- per$frsbosdat <- NULL
per$fid <- per$fboort <- per$ab <- per$fyrkfd <- per$fyrkind <- NULL
per$region <- per$kon <- per$fodhfrs <- per$frsbobdat <- per$frsbostyp <- NULL
per$fbonmn <- per$fboind <- per$fyrkrel <- per$fyrketxt <- NULL
per$foddat <- as.Date(as.character(per$foddat), format = "%Y%m%d")
## NEW:
per$socBranch <- as.character(per$socBranch)
per$socBranch[is.na(per$socBranch)] <- "none"
per$socBranch <- factor(per$socBranch)
per$socStatus <- as.character(per$socStatus)
per$socStatus[is.na(per$socStatus)] <- "low"
per$socStatus <- factor(per$socStatus)
per <- per[, c("id", "sex", "birthdate", "mid", "foddat", "parity", "illeg", 
               "subreg", "socBranch", "socStatus", "hisclass", "enter", "exit", "event")]
#saveRDS(per, file = "data/infdat3.rds")
```

### 2022-02-08 paper4.Rmd
fix errors and remove socStatus

---
title: "A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950"
author: "Göran Broström and Tommy Bengtsson"
date: "`r Sys.time()`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
  bookdown::word_document2:
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
citation_package: natbib
classoption: titlepage
bibliography: bio.bib
titlepage: yes
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{a4wide}
#- \usepackage[figuresonly]{endfloat}
#- \renewcommand{\efloatseparator}{\mbox{}}
abstract: "The effect of extreme temperatures on infant mortality in the Umeå and Skellefteå regions 1895-1950 is studied in a biometric analysis setting. More precisely, the effect of climate and weather, measured by temperature, average and extremes, on infant mortality is investigated. It turns out that climate (average) is more important than weather (extremes), low average temperatures are more important than temporary dips in temperature."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=3, comment=NA)
knitr::opts_knit$set(global.par = TRUE)
##
library(knitr)
library(eha)
```

@@

```{r prepare}
source("R/ltx2.R")
infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
##infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951), dig.lab = 5)
##levels(infdat$subreg) <- c("ume", "ske")
##infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
##infdat$season <- as.factor(infdat$season)
## New:
infdat$summer <- infdat$week %in% 14:39
infdat$emeantemp <- round(infdat$emeantemp)
levels(infdat$socBranch) <- c("office", "farming", "office", "worker", "worker") # NOTE!
```

@@

We separate the investigation into two parts, *neonatal* and *postneonatal* mortality. But first,
a joint analysis.

```{r joint, cache = TRUE}
library(survival)
neo <- age.window(infdat, c(0, 28 / 365))
neo$ageIvl <- "neo"
pneo <- age.window(infdat, c(28/365, 1))
pneo$ageIvl <- "postneo"
all <- rbind(neo, pneo)
all$ageIvl <- factor(all$ageIvl)
##all$enter <- round(all$enter * 365)
##all$exit <- round(all$exit * 365)
fit.all <- coxph(Surv(enter, exit, event) ~ strata(ageIvl) * (aver + emeantemp + lowTemp + highTemp + socBranch + socStatus + sex + period),
                 data = all)
dr.all <- drop1(fit.all, test = "Chisq")

inftab <- toTpch(Surv(enter, exit, event) ~ subreg + sex + birthmonth + period + socBranch + socStatus + parity + 
                     aver + emeantemp + lowTemp + highTemp + summer + ageIvl, data = all,
                 cuts = c(0, 3, 7, 14, 28, 56, 84, 145, 207, 286, 365) / 365)
tabneo <- inftab[inftab$ageIvl == "neo", ]
tabpost <- inftab[inftab$ageIvl == "postneo", ]
tabpost$age <- factor(tabpost$age)
tabneo$age <- factor(tabneo$age)
##system.time(tabneo.fit <- tpchreg(oe(event, exposure) ~ aver + emeantemp + highTemp + lowTemp + socBranch + socStatus, 
  ##               data = tabneo, time = age))
##system.time(tabpost.fit <- tpchreg(oe(event, exposure) ~ aver + emeantemp + highTemp + lowTemp + socBranch + socStatus, 
  ##               data = tabpost, time = age))
```

```{r shownonprop}
tbl(dr.all, caption = "Proportionality tests, neonatal vs. postneonatal.", rownames = TRUE)
```

@@

@@

```{r neoanalw, results = 'asis'}
##library(survival)
winter <- neo[!neo$summer, ]
##winter$season <- factor(winter$season)
##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
##neo$enter <- g(neo$enter * 365)
fit.cr <- coxreg(Surv(enter, exit, event) ~  highTemp + aver + emeantemp + 
                     sex + subreg + illeg + parity + socBranch + period,
                 data = winter)
dr <- drop1(fit.cr, test = "Chisq")
ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, October to March. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
     label = "tab:neoanalw", keep = NULL)
```

@@

```{r neoanals, results = 'asis'}
##library(survival)
#summer <- neo[neo$summer, ]
##summer$season <- factor(summer$season)
##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
##neo$enter <- g(neo$enter * 365)
fit.cr <- tpchreg(oe(event, exposure) ~  aver + emeantemp + 
                     sex + subreg  + socBranch + period,
                 data = inftab[inftab$summer, ], time = age)
dr <- drop1(fit.cr, test = "Chisq")
ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, April to September. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
     label = "tab:neoanals", keep = NULL)
```

@@

fit.tpch <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch  + 
                        parity + subreg + period + summer, data = tabpost[!tabpost$summer, ], time = age)
dr <- drop1(fit.tpch, test = "Chisq")
ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, October to March. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanalw", keep = NULL)
```

The result in Table \@ref(tab:postanalw) shows that *climate* (`emintemp`) is more important than 
*weather* (`excessTemp`). Moreover, no signs of interaction between weather or climate and the rest of 
covariates (not shown).

>>>>

---
title: "A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950"
author: "Göran Broström and Tommy Bengtsson"
date: "`r Sys.time()`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
  bookdown::word_document2:
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
citation_package: natbib
classoption: titlepage
bibliography: bio.bib
titlepage: yes
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{a4wide}
- \usepackage[tablesonly]{endfloat}
- \renewcommand{\efloatseparator}{\mbox{}}
abstract: "The effect of extreme temperatures on infant mortality in the Umeå and Skellefteå regions 1895-1950 is studied in a biometric analysis setting. More precisely, the effect of climate and weather, measured by temperature, average and extremes, on infant mortality is investigated. It turns out that climate (average) is more important than weather (extremes), low average temperatures are more important than temporary dips in temperature."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=3, comment=NA, cache= TRUE) ## NOTE!!
knitr::opts_knit$set(global.par = TRUE)
##
library(knitr)
library(eha)
```

@@

```{r prepare}
source("R/ltx2.R")
infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
##infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951), dig.lab = 5)
##levels(infdat$subreg) <- c("ume", "ske")
##infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
##infdat$season <- as.factor(infdat$season)
## New:
infdat$summer <- infdat$week %in% 14:39
infdat$emeantemp <- round(infdat$emeantemp)
levels(infdat$socBranch) <- c("office", "farming", "none", "office", "worker") # NOTE!
```

@@

Comments on other candidate covariates:

*Birth month* is left out in the analyses despite that fact that
it is an important factor in neonatal mortality. However, we include *time of year* in terms of 
*winter* and *summer* as a time-varying covariate, and in the neonatal case it will coincide to a
great extent with birth period. For the postneonal case the situation is different, but it turns out that
for those infants who have survived the first month of life, birth month does not matter much.
We separate the investigation into two parts, *neonatal* and *postneonatal* mortality. But first,
a joint analysis.

*Socioeconomic status* is divided into two factor covariates: *socBranch* and *socStatus*. The latter should be 
seen relative to actual *socBranch*. We have information on whether the infant was *illegitimate* (mother unmarried),
but we incorporate those cases in the category *none* of *socBranch* and *unknown* as category of *socStatus*.

```{r joint, cache = TRUE}
library(survival)
neo <- age.window(infdat, c(0, 28 / 365))
neo$ageIvl <- "neo"
pneo <- age.window(infdat, c(28/365, 1))
pneo$ageIvl <- "postneo"
all <- rbind(neo, pneo)
all$ageIvl <- factor(all$ageIvl)
##all$enter <- round(all$enter * 365)
##all$exit <- round(all$exit * 365)
##fit.all <- coxph(Surv(enter, exit, event) ~ strata(ageIvl) * (aver + emeantemp + lowTemp + highTemp + socBranch + ##socStatus + sex + period),
  ##               data = all)
##dr.all <- drop1(fit.all, test = "Chisq")

inftab <- toTpch(Surv(enter, exit, event) ~ subreg + sex + period + socBranch + socStatus + parity + 
                     aver + emeantemp + lowTemp + highTemp + summer + ageIvl, data = all,
                 cuts = c(0, 3, 7, 14, 28, 56, 84, 145, 207, 286, 365) / 365)
tabneo <- inftab[inftab$ageIvl == "neo", ]
tabpost <- inftab[inftab$ageIvl == "postneo", ]
tabpost$age <- factor(tabpost$age)
tabneo$age <- factor(tabneo$age)
fit.all <- tpchreg(oe(event, exposure) ~  strata(ageIvl) * (aver + emeantemp + lowTemp + highTemp + socBranch + socStatus + sex + period) + emeantemp, data = inftab, time = age)
dr.all <- drop1(fit.all, test = "Chisq")
##system.time(tabneo.fit <- tpchreg(oe(event, exposure) ~ aver + emeantemp + highTemp + lowTemp + socBranch + socStatus, 
  ##               data = tabneo, time = age))
##system.time(tabpost.fit <- tpchreg(oe(event, exposure) ~ aver + emeantemp + highTemp + lowTemp + socBranch + socStatus, 
  ##               data = tabpost, time = age))
```

```{r shownonprop}
tbl(round(dr.all, 4), caption = "Proportionality tests, neonatal vs. postneonatal.", rownames = TRUE)
```

From Table \@ref(tab:shownonprop) we can see that the variables *emeantemp*, *lowTemp*, *socStatus* and *period* all have different effects on mortality in the neonatal case compared to the postneonatal case. In terms of 
*infant mortality*, we can say that these caovariates have non-proportional effects. As an example, see Figure

@@

```{r formneo}
formneo <- as.formula("oe(event, exposure) ~  lowTemp + highTemp + aver + emeantemp + 
                     sex + subreg + parity + socStatus + period")
```

@@

```{r neoanalw, results = 'asis'}
##library(survival)
winter <- tabneo[!tabneo$summer, ]
##winter$season <- factor(winter$season)
##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
##neo$enter <- g(neo$enter * 365)
fit.cr <- tpchreg(formneo, data = winter, time = age)
dr <- drop1(fit.cr, test = "Chisq")
ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, October to March. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
     label = "tab:neoanalw", keep = NULL)
```

@@

```{r neoanals, results = 'asis'}
##library(survival)
#summer <- neo[neo$summer, ]
##summer$season <- factor(summer$season)
##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
##neo$enter <- g(neo$enter * 365)
fit.cr <- tpchreg(formneo, data = tabneo[tabneo$summer, ], time = age)
dr <- drop1(fit.cr, test = "Chisq")
ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, April to September. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
     label = "tab:neoanals", keep = NULL)
```

@@

fit.tpch <- tpchreg(oe(event, exposure) ~ period * (lowTemp + highTemp + aver + emeantemp) + sex + socBranch  + parity + subreg + period, data = tabpost[!tabpost$summer, ], time = age)
dr <- drop1(fit.tpch, test = "Chisq")
tbl(round(dr, 4), caption = "Interaction tests with period, postneonatal mortality.", rownames = TRUE)
##ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, October to March. Adjusted for social branch, sex, ##illegitimacy, parity, time period, and subregion.", label = "tab:postanalw", keep = NULL)
```

The result in Table \@ref(tab:postanalw) shows that the temperature effects vary with time period,
so we continue with seaparate analyses for the time periods `r levels(tabpost$period)`.

```{r setformw}
form <- as.formula("oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socStatus + parity + subreg")
```

#### The time period 1895--1913

```{r postw1, results = "asis"}
fit.pw1 <- tpchreg(form, data = tabpost[(!tabpost$summer) & tabpost$period == "(1895,1914]", ], time = age)
dr <- drop1(fit.pw1, test = "Chisq")
ltx2(fit.pw1, dr = dr, caption = "Postneonatal mortality, October to March 1895-1914. Adjusted for social branch, sex, illegitimacy, parity, and subregion.", label = "tab:postw1", keep = NULL)
```

Table \@ref(tab:postw1) shows that

#### The time period 1914--1934

```{r postw2, results = "asis"}
fit.pw2 <- tpchreg(form, data = tabpost[(!tabpost$summer) & tabpost$period == "(1914,1935]", ], time = age)
dr2 <- drop1(fit.pw2, test = "Chisq")
ltx2(fit.pw2, dr = dr2, caption = "Postneonatal mortality, October to March 1914-1934. Adjusted for social branch, sex, illegitimacy, parity, and subregion.", label = "tab:postw2", keep = NULL)
```

Table \@ref(tab:postw2) shows that

#### The time period 1935--1950

```{r postw3, results = "asis"}
fit.pw3 <- tpchreg(form, data = tabpost[(!tabpost$summer) & tabpost$period == "(1935,1951]", ], time = age)
dr3 <- drop1(fit.pw3, test = "Chisq")
ltx2(fit.pw3, dr = dr3, caption = "Postneonatal mortality, October to March 1935-1950. Adjusted for social branch, sex, illegitimacy, parity, and subregion.", label = "tab:postw3", keep = NULL)
```

Table \@ref(tab:postw3) shows that
